Gravity disturbance or gravity anomaly?

### 2018-04-16 manuscript/letter.tex

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called simply 
\textit{gravity} \citep{hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.
%In applied geophysics, gravity surveys are usually conducted 
%over local or regional areas on the Earth's surface for the purpose of 
%characterizing geological structures located within the
%crust and upper mantle.
%
%Consequently, geophysicists are only interested in the
%particular gravitational component of gravity which 
%is produced by these target geological structures.
The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics 
\citep{blakely1996}.

@@

The normal Earth gives rise to the \textit{geodetic coordinate system}
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

At a given point $(h, \varphi, \lambda)$, there are three 
unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: ,
\label{eq:unit-vectors}
\end{equation}
Notice that these unit vectors are mutually orthogonal.
The normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.

@@

It is worth noting that, by definition, 
the centrifugal component of the normal gravity field is
equal to the centrifugal component of the Earth's gravity
field if they are evaluated at the same point.
Then, the differences between the gravity vector
(corrected from non-gravitational effects) 
and the normal gravity vector, at the same point, represents a purely 
gravitational and consequently harmonic disturbing field.
This gravitational disturbance 
is caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
\citeyear{blakely1996}). Here, we opted for using the last term.

The difference between the observed gravity and the
normal gravity, at the same point, is called gravity disturbance
\citep{hofmann-wellenhof-moritz2005}.
Notice that the gravity disturbance is not equivalent to the
magnitude of the difference between the gravity vector
and the normal gravity vector at the same point.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

@@

By definition, the gravity disturbance $\delta g_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by 
\citep{hofmann-wellenhof-moritz2005}:
\begin{equation}
\delta g_{i} =  g_{i} - \gamma_{i} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{i} = \| \mitbf{g}_{i} \|$ and 
$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition 
$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
$g_{i}$ by a first order Taylor's expansion as follows 
\citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$ and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(equation \ref{eq:unit-vectors}) evaluated at the point $(h_{i}, \varphi_{i}, \lambda_{i})$, which is obtained by transforming $(x_{i}, y_{i}, z_{i})$
from the topocentric coordinate system (Figure \ref{fig:fig2}) to the
geodetic coordinate system (Figure \ref{fig:fig1}).

PAREI AQUI

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
coincides with the $z$ axis
of the local Cartesian coordinate system defined
at the beginning of this section. Consequently,
by using the approximation defined in equation \ref{eq:gobs-approx}, 
the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
rewritten as follows \begin{equation}
\delta g^{o}_{i} \approx \hat{\mitbf{z}}^{\top} \Delta \mitbf{g}^{o}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
According to equation \ref{eq:gravity-disturbance-approx},
the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
represents the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source can be represented 
by the following harmonic function
\begin{equation}
d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

\section{Gravity disturbance versus gravity anomaly}

Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms).

Consequently, they implicitly or directly assume that
the gravity data approximates the gravitational disturbance.

Almost all forward modelling techniques compute
the vertical component of the gravitational attraction 
exerted by the geological bodies at the observation points.

Hence, almost all geophysicists implicitly compute 
the vertical component of the gravitational disturbance.

The gravity anomaly is defined as the difference between 
the gravity $\| \mitbf{g}_{P} \|$, at a point $P$ on the
Geoid, and the normal gravity $\| \mitbf{\gamma}_{Q} \|$,
on the reference elipsoid, where $P$ and $Q$ have the same
geodetic latitude and longitude.

Consequently, the gravity anomaly is a function of
the geodetic latitude and longitude only and cannot
be calculated at arbitrary heights.

Gravity anomalies require the computation of gravity
within the topographic masses, where the gravitational
disturbance is not harmonic.

>>>>

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called simply 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.
%In applied geophysics, gravity surveys are usually conducted 
%over local or regional areas on the Earth's surface for the purpose of 
%characterizing geological structures located within the
%crust and upper mantle.
%
%Consequently, geophysicists are only interested in the
%particular gravitational component of gravity which 
%is produced by these target geological structures.
The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics 
\citep{blakely1996}.

@@

The normal Earth gives rise to the \textit{geodetic coordinate system}
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).
At a given point $(h, \varphi, \lambda)$, there are three 
unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: ,
\label{eq:unit-vectors}
\end{equation}
Notice that these unit vectors are mutually orthogonal and
that the normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.

@@

It is worth noting that, by definition, 
the centrifugal component of the normal gravity field is
equal to the centrifugal component of the Earth's gravity
field if they are evaluated at the same point.
Then, the differences between the gravity vector
(corrected from non-gravitational effects) 
and the normal gravity vector, at the same point, represents a purely 
gravitational and consequently harmonic disturbing field.
This gravitational disturbance 
is caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
or \textit{gravity sources} (e.g., \citeauthor{blakely1996}, 
\citeyear{blakely1996}). Here, we opted for using the last term.

The difference between the observed gravity and the
normal gravity, at the same point, is called \textit{gravity disturbance}
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
Notice that the gravity disturbance is not equivalent to the
magnitude of the difference between the gravity vector
and the normal gravity vector at the same point 
\citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

@@

For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
Cartesian coordinate system (Figure \ref{fig:fig2}b), there is
a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
in the geodetic coordinate system (Figure \ref{fig:fig1}).
Here, we opted for omitting the equations to convert a point
$(x_{i}, y_{i}, z_{i})$ into a point $(h_{i}, \varphi_{i}, \lambda_{i})$
and vice versa. These equations, however, may be easily
found in the literature \citep[e.g.,][]{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.

By definition, the gravity disturbance $\delta g_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:
\begin{equation}
\delta g_{i} =  g_{i} - \gamma_{i} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{i} = \| \mitbf{g}_{i} \|$ and 
$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition 
$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
$g_{i}$ by a first order Taylor's expansion as follows 
\citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Figure \ref{fig:fig2}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(equation \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
geodetic coordinate system (Figure \ref{fig:fig1}).

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the topocentric Cartesian coordinate 
system (Figure \ref{fig:fig2}b).
Consequently, by using the approximation defined in equation 
\ref{eq:gobs-approx}, the gravity disturbance 
(equation \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the topocentric Cartesian coordinate system
(Figure \ref{fig:fig2}).
This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector. In the topocentric Cartesian coordinate system,
the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
can be defined as the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source with density contrast $\Delta\rho$
(in $kg / m^{3}$) can be represented by the following harmonic function:
\begin{equation}
d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

Practically all the literature about gravity modeling
use the quantity $d_{i}$ (equation \ref{eq:gz-local})
to represent the gravity anomaly produced by 
gravity sources \citep[e.g.,][]{blakely1996}. 
Consequently, almost all the geophysicists use the gravity
anomaly as an approximation of the gravity disturbance
produced by the gravity sources. Notice that $d_{i}$ 
(equation \ref{eq:gz-local}) does not depend on the
geoidal surface. Rather, it depends on the relative position 
of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
the gravity sources in a topocentric Cartesian coordinate
system (Figure \ref{fig:fig2}b).

\section{Conclusions}

We debate the conceptual differences between the gravity
disturbance and the gravity anomaly.
Our reasoning suggest that the gravity disturbance is the
more appropriated quantity for representing the gravity
effect produced by the gravity sources.
In summary, we point out that:

\begin{enumerate}
%\renewcommand{\theenumi}{(\arabic{enumi})}
\renewcommand{\theenumii}{\alph{enumii}}

\item Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms). As a consequence, they implicitly or directly assume that
the gravity data approximates the gravitational disturbance.

\item Almost all forward modeling techniques compute
the vertical component of the gravitational attraction 
exerted by the geological bodies at the observation points.
Notice that the gravity anomaly requires the computation of gravity
on the Geoid, which is generally within the topographic masses.
Hence, almost all geophysicists implicitly compute 
the gravity disturbance.

\item The gravity anomaly is defined as the difference between 
the gravity on the Geoid and the normal gravity on the reference 
ellipsoid, both at the same geodetic latitude and longitude.
Consequently, the gravity anomaly is a function of
the geodetic latitude and longitude only and cannot
be calculated at arbitrary heights. On the other hand,
the gravity disturbance can be computed at arbitrary points
outside the sources.

\end{enumerate}

The principal theoretical implication of this study is that
is that, although the gravity anomaly may be used as a good approximation of
the gravity effect produced by the sources for most practical applications,
the more appropriated quantity for gravity modeling is the
gravity disturbance. We stress that, more important than the 
the terminology changes, the geophysicist must bear in mind the 
conceptual assumptions used in gravity modeling.
